Sparse Function-space Representation \\ of Neural Networks

### 2023-05-15 paper/main.tex
Section titles
updates cl part

\section{Introduction}
\label{sec:intro}
%
Deep learning \cite{goodfellow2016deep} has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly, how can we efficiently quantify uncertainty without resorting to expensive and hard-to-interpret sampling in the model's weight-space? Secondly, how to update the weights of an already trained model with new batches of data without compromising the performance on past data? These questions become central when applied to sequential learning paradigms, such as continual learning (CL, \cite{de2021continual}) and reinforcement learning (RL, \cite{sutton2018reinforcement}). In CL, access to the previous data is lost and the challenge is retaining a compact representation of the problem and avoiding forgetting over the life-long learning horizon~\cite{mccloskey1989catastrophic}. Similarly, in RL, the model must adapt to environmental observations through exploration, while leveraging prediction uncertainties to assess potential future paths.\looseness-1

@@

%Current state of affairs 
Recent techniques (\eg, \cite{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable}) for converting trained neural networks from weight-space to function-space allow a principled mathematical approach for analyzing the behaviour \cite{cho2009kernel,meronen2020stationary}, performing probabilistic inference \cite{khan2019approximate}, and quantifying uncertainty in neural networks \cite{foong2019between}. These methods rely on the linearization of neural networks\cite{jacot2018neural} the neural network and capturing the first two moments in function space in terms of a mean function and covariance function (or kernel)---defining a Gaussian process (GP, \cite{rasmussen2006gaussian}). GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates, and they serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}. 
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
In this paper, we demonstrate that any neural network can be represented in terms of the so-called `dual' parameters of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast}. In contrast to previous work that utilizes subsets \cite{immer2021scalable}, this parameterization allows for the consolidation of contributions from {\em all} training data into a sparse representation. Crucially, the resulting GP now predicts in the same space as the originally trained neural network, a feature not present in previous approaches, while still avoiding the notorious cubic complexity of vanilla GPs. Through the dual parameter formulation, we establish a connection between the neural network, the resulting full GP, and a sparse approximation similar to variational sparse GPs~\cite{titsias2009variational}. Using the sparse GP we can then take advantage of the property of dual parameters that allows for fast conditioning recently shown in \cite{chang2022fantasizing}. Thus we are able to avoid retraining and condition new data in to our model. The resulting \our method can be applied in any sequential decision-making scenario where retraining is not feasible.

@@

The contributions of this paper are:
%
{\em (i)}~We introduce \our, a new approach for sparse representation of the neural network in the function space.
{\em (ii)}~We demonstrate that, despite its sparsity, our method effectively captures predictive uncertainty, provides means of updating the model post-training, and gives a a compact representation suitable for continual learning.
{\em (iii)}~We provide extensive experiments for showcasing our approach and demonstrate significance and applicability across supervised, continual, and reinforcement learning, aiming to stimulate future use of the approach.

@@

\textbf{Function-space methods for sequential learning} The central problem in continual learning is the non-stationary nature of the training distribution which causes the training to overwrite the previously learnt parameters---leading to the model forgetting the previous learnt functions. This problem is typically tackled in weight-space with regularisation methods such as EWC, \cite{kirkpatrick2017overcoming}, SI~\cite{zenke17a}), or VCL~\cite{nguyen-tuongModel2009} that induce retention of previously learned information in the weights. However, encoding functional knowledge in the weights directly is challening, and thus recent methods such as DER~\cite{buzzega2020dark}, FROMP~\cite{pan2020continual}, and S-FSVI~\cite{rudner2022continual} have focused on function-space regularization.

%% CL section 
%%
%% EWC, SI, VCL -> weight regularization 
% 1) what is the central problem of CL? 
% 		Training from a non-stationary distribution the models overwrites the previously learnt parameters and then forgets the previous learnt function.
% 2) What others have done?
% 		starting with EWC and SI, that are the most known methods in term of weight regularization. And others took a different approach again in the weight regularization real with VCL (to check)
% 3) People started to try experimenting with the functional regularization -> 
% In order of complexity -> DER, FROMP, S-FSVI
% In order of pubblication -> DER/FROMP, K-Priors (?), S-FSVI
%EWC \cite{kirkpatrick2017overcoming}
%SI \cite{zenke17a}, 
%VCL \cite{nguyen-tuongModel2009},

%% DER, FROMP, S-FSVI
%DER \cite{buzzega2020dark}
%FROMP \cite{pan2020continual}
%S-FSVI \cite{rudner2022continual}

@@

\textbf{Uncertainty quantification}
Bayesian inference offers a principled approach to quantifying uncertainty in neural networks.
The goal is to find the posterior over the weights ${p(\vw \mid \vy) \propto p(\vy \mid f_{\weights}(\vx)) p(\weights)}$ as
it represents our belief in the parameters after combining data $\dataset$ with our prior $p(\vw)$.
Although the true posterior $p(\vw \mid \dataset)$ is intractable given the non-linearities of the neural network, one can resort to sampling techniques such as Hamiltonian Monte-Carlo (HMC).
However, for most applications their high computational costs make them impractical so we need approximate inference techniques.
In \cref{fig:banana}, we show a qualitative example of the induced function-space posterior obtained with our method compared to the posterior obtained through HMC sampling on the Banana toy dataset. \todo{make it more straightforward that we are computationally cheaper? A: Not sure we are reallying trying to sell our method here}

@@

\subsection{Continual Learning}
% Definition of CL -> a sequence of tasks

@@

    % THE TABLE NUMBER ARE GENERATED BY A SCRIPT	
	\input{tables/img_super.tex}
\end{table}

Similarly to the UCI experiments, we trained neural networks (CNN, architexture matches \citet{immerImprovingPredictionsBayesian2021}) for the FMNIST and CIFAR10 classification tasks. After training, we computed the duals for NN2SVGP and used the resulting SVGP for uncertainty estimates, see \cref{tbl:imagesuper}. We performed hyperparameter optimization over the prior precision, and found the optimal value to be [insert]. We ran the experiment over $5$ seeds. During the NN training, we used a batch size of $512$ and an Adam optimizer with learning rate $1e-3$.

In addition to uncertainty estimates, our method can update the posterior when given new data. We gave the SVGP model 10 \% of the test dataset used for evaluation, and compare the performance of the SVGP model and the retrained NN in [INSERT TABLE]. 
\subsection{Updating the representation during continual learning}

\begin{table}[t!] 
  \centering\scriptsize
  \caption{
  TODO: CL Experiments. $^*$ Methods relying only on weight regularization. 
  \color{blue}{Results from \cite{rudner2022continual}.} 
  %\color{red}{Preliminary results need more runs or hyperparam tuning}
  }
	\label{tbl:cl_table_1}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{2pt}
	\setlength{\tblw}{0.14\textwidth}

>>>>

\section{Introduction}
\label{sec:intro}
%
Deep learning \cite{goodfellow2016deep} has become the cornerstone of contemporary artificial intelligence, proving remarkably effective in tackling supervised and unsupervised learning tasks in the {\em large data}, {\em offline}, and {\em gradient-based training} regime. Despite its success, gradient-based learning techniques exhibit limitations. Firstly, how can we efficiently quantify uncertainty without resorting to expensive and hard-to-interpret sampling in the model's weight-space? Secondly, how to update the weights of an already trained model with new batches of data without compromising the performance on past data? These questions become central when applied to sequential learning paradigms, such as continual learning (CL, \citep{parisi2019continual, de2021continual}) and reinforcement learning (RL, \cite{sutton2018reinforcement}). In CL, access to the previous data is lost and the challenge is retaining a compact representation of the problem and avoiding forgetting over the life-long learning horizon~\cite{mccloskey1989catastrophic}. Similarly, in RL, the model must adapt to environmental observations through exploration, while leveraging prediction uncertainties to assess potential future paths.\looseness-1

@@

%Current state of affairs 
Recent techniques (\eg, \cite{ritter2018kfac,khan2019approximate,daxberger2021laplace,fortuin2021bayesian,immer2021scalable}) for converting trained neural networks from weight-space to function-space allow a principled mathematical approach for analyzing the behaviour \cite{cho2009kernel,meronen2020stationary}, performing probabilistic inference \cite{khan2019approximate}, and quantifying uncertainty in neural networks \cite{add}. These methods rely on the linearization of neural networks\cite{jacot2018neural} the neural network and capturing the first two moments in function space in terms of a mean function and covariance function (or kernel)---defining a Gaussian process (GP, \cite{rasmussen2006gaussian}). GPs provide a widely-used probabilistic toolkit with principled uncertainty estimates, and they serve as a standard surrogate model for Bayesian optimization \citep{garnett_bayesoptbook_2022} and are effective in model-based reinforcement learning \citep{deisenroth2011pilco} with theoretical guarantees on regret bounds \citep{srinivas2009gaussian}. 
%Yet many problems lie in high dimensional input space; for example, images are where GPs cannot learn representations. In such scenarios such as in many reinforcement learning environments neural networks are used as the surrogate model. However, uncertainty is still essential to ensure effective exploration for sequential algorithms. Successful approaches have attempted to blend neural networks with uncertainty estimates around predictions, allowing for sophisticated exploration strategies. However, there has been limited use of hybrid models that possess the feature representation ability of neural networks but also attractive the properties of GPs, such a hybrid method we propose in this paper.

%Need for adaptive learning methods + failures with current methods
In this paper, we demonstrate that any neural network can be represented in terms of the so-called `dual' parameters of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast}. Contrasting to previous work that use subsets \cite{immer2021scalable}, this parameterization allows squashing the contributions of {\em all} training data into a sparse representation. Crucially, the resulting GP now predicts in the same space as the originally trained neural network, a feature not present in previous approaches, while still avoiding the notorious cubic complexity of vanilla GPs. Given the dual parameter formulation, we establish a link between the neural network, the resulting full GP, and a sparse approximation of it akin to variational sparse GPs~\cite{titsias2009variational}. Using the sparse GP we can then take advantage of the property of dual parameters that allows for fast conditioning recently shown in \cite{chang2022fantasizing}. Thus we are able to avoid retraining and condition new data in to our model. The resultant \our method can be used in any sequential decision-making scenario where retraining is not available.
%\todo{a pragraph on how this can be used for improving CL and RL}

@@

The contributions of this paper are as follows.
%
{\em (i)}~We introduce \our, a new approah for sparse representation of the neural network in the function space.
{\em (ii)}~We show that, despite being sparse, the method gives a good   
A compact representation for continual learning
{\em (iii)}~Fast updates for adaptive exploration.
%
We provide extenive experiments for showcasing our approach and demonstrate importance and applicability across supervised, continual, and reinforcement learning to stimulate future use of the approach.

@@

% CL section
\textbf{Function-space methods for sequential learning} The central problem in continual learning is how to deal with the non-stationary nature of the training distribution which causes the training process to overwrite the previously learnt parameters---leading to the model forgetting the previous learnt functions. 
The approaches proposed in the CL literature can be categorized into inference-based, reharsal-based, and model-based methods, we refer the reader to \citep{parisi2019continual, de2021continual} for complete reviews. Many methods in the first category tackled this problem with weight-space regularization, such as EWC~\cite{kirkpatrick2017overcoming}, SI~\cite{zenke17a}), or VCL~\cite{nguyen-tuongModel2009} that induce retention of previously learned information in the weights alone.
However, encoding functional knowledge in the weights does not guarantees a good quality of the predictions, and thus recent methods such as DER~\cite{buzzega2020dark}, FROMP~\cite{pan2020continual}, and S-FSVI~\cite{rudner2022continual} have focused on function-space regularization. 
% It should be written clearly that our method actually is doing a "functional" reharsal, because we store a set of inducing points that are replayed in order to check if the logits change.

% Other methods that could be worth citing are 
% Learning without forgetting \citep{li2018lwf} :  first paper proposing to use functional regularization computes a smoothed version of the current logits for the new examples at the beginning of each task, minimizing their drift during training. It doesn't have any set of inducing points
% \citep{benjamin2018measuring} this is not a proper CL paper, but it's usually referred in the functional regularization papers, because shows how to improve Adam with measuring the network difference in function space. They also have some easy CL examples there.

@@

\textbf{Uncertainty quantification}
Bayesian inference offers a principled approach to quantifying uncertainty in neural networks.
The goal is to find the posterior over the weights ${p(\vw \mid \vy) \propto p(\vy \mid f_{\weights}(\vx)) p(\weights)}$ as
it represents our belief in the parameters after combining data $\dataset$ with our prior $p(\vw)$.
Although the true posterior $p(\vw \mid \dataset)$ is intractable given the non-linearities of the neural network, one can resort to sampling techniques such as Hamiltonian Monte-Carlo (HMC).
However, for most applications their high computational costs make them impractical so we need approximate inference techniques.
In \cref{fig:banana}, we show a qualitative example of the induced function-space posterior obtained with our method compared to the posterior obtained through HMC sampling on the Banana toy dataset.

@@

\subsection{Continual Learning}
% Definition of CL -> a sequence of tasks 
In this paper we consider the Continual learning setting were a sequence of disjoint sets

@@

    % THE TABLE NUMBER ARE GENERATED BY A SCRIPT	
	\input{tables/img_super.tex}
\end{table}
\subsection{Updating the  network representation for Continual learning}
The Continual learning setting is a challenging and the training and
\begin{table}[t!] 
  \centering\scriptsize
  \caption{
  TODO: CL Experiments. $^*$ Methods relying only on weight regularization. 
  \color{blue}{Results from \cite{rudner2022continual}.} 
  %\color{red}{Preliminary results need more runs or hyperparam tuning}
  }
	\label{tbl:cl_table_1}
	% Control table spacing
	\renewcommand{\arraystretch}{1.}
	\setlength{\tabcolsep}{2pt}
	\setlength{\tblw}{0.14\textwidth}
